Generating function solution for bursty promoters

### 2020-04-20 doc/gen_fcn_notes.tex
Added more to writeup on gen fcn + numerics for 2F1

@@

just consider mRNA \mmnote{replicating Charlotte's notes; she says it's worked out

@@

comprehend}. The master equation of interest is

@@

\section{Adding repression}
\subsection{Deriving the generating function for mRNA distribution}
Let us move from a one-state promoter to a two-state promoter, where one state
has repressor bound and the other produces transcriptional bursts as above.
Although now we have an equation for each promoter state, otherwise
the master equation reads similarly to the one-state case, except with
additional terms corresponding to transitions between promoter states, namely
\begin{align}
\deriv{t}p_R(m,t) =& k_R^+ p_A(m,t) - k_R^- p_R(m,t)
        + (m+1)\gamma p_R(m+1,t) - m\gamma p_R(m,t)
\\
\begin{split}
\deriv{t}p_A(m,t) =& - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
        + (m+1)\gamma p_A(m+1,t) - m\gamma p_A(m,t) 
\\
&- k_i p_A(m,t) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime,t),
\end{split}
\end{align}
where $p_R(m,t)$ is the probability of the system having $m$ mRNA copies
and having repressor bound to the promoter at time $t$, and $p_A$ is an
analogous probability to find the promoter without repressor bound.
$k_R+$ and $k_R^-$ are, respectively, the rates at which repressors
bind and unbind to and from the promoter, and
$\gamma$ is the mRNA degradation rate.
$k_i$ is the rate at which bursts initiate, and as before,
the geometric distribution of burst sizes has mean $b=(1-\theta)/\theta$.\footnote{
$\theta$ can be viewed as a Bernoulli trial probability: after each transcript
in a burst is produced, $\theta$ is the probability that the burst
terminates and $(1-\theta)$ is the probability that the burst continues.
}

@@

any standard source, e.g.~\cite{Abramovitz}, and identify the

@@

\subsection{Numerical considerations and recursion formulas}
We would like to carry out Bayesian parameter inference on FISH data
from~\cite{Jones2014}, using~\eref{eq:p_m_bursty+rep} as our
likelihood. This requires accurate (and preferably fast)
numerical evaluation of the hypergeometric function ${_2F_1}$,
which is a notoriously hard problem~\cite{Pearson2016, Gil2007},
and our particular needs here present an especial challenge as we show below.

@@

The solution involves recursion relations originally explored by Gauss,
and studied extensively in~\cite{Pearson2016, Gil2007}.

@@

% \bibliographystyle{nature}
\bibliographystyle{abbrv}
\bibliography{../library}

>>>>

\begin{figure}
\includegraphics[width=0.9\textwidth]{../figures/fig2/fig2.pdf}
\caption{
\mmnote{This figure wouldn't appear in this section, instead
just a reference to the cartoon so the reader knows
what cartoon model we're doing all this math for.}
\end{figure}
We would like to derive the full probability distribution of
mRNA copy number per cell for a model combining simple repression
with a promoter featuring bursty transcription initiation.

@@

just consider the bursty consititutive promoter,
model 4 in Fig.~1A.
\mmnote{replicating Charlotte's notes; she says it's worked out

@@

comprehend. Need to find a cite for this,
or maybe just cite Charlotte, personal comm, since that's where I actually got it\dots}.
The master equation of interest is

@@

\section{Adding repression}
\subsection{Deriving the generating function for mRNA distribution}

\begin{figure}
\includegraphics[width=0.9\textwidth]{../figures/fig3/fig3.ai}
\caption{
\mmnote{This figure wouldn't appear in this section, instead
just a reference to the cartoon so the reader knows
what cartoon model we're doing all this math for.}
}
\end{figure}
Let us move from a one-state promoter to a two-state promoter, where one state
has repressor bound and the other produces transcriptional bursts as above.
A schematic of this model is shown in Fig.\ 2A.
Although now we have an equation for each promoter state, otherwise
the master equation reads similarly to the one-state case, except with
additional terms corresponding to transitions between promoter states, namely
\begin{align}
\deriv{t}p_R(m,t) =& k_R^+ p_A(m,t) - k_R^- p_R(m,t)
        + (m+1)\gamma p_R(m+1,t) - m\gamma p_R(m,t)
\\
\begin{split}
\deriv{t}p_A(m,t) =& - k_R^+ p_A(m,t) + k_R^- p_R(m,t)
        + (m+1)\gamma p_A(m+1,t) - m\gamma p_A(m,t) 
\\
&- k_i p_A(m,t) + k_i \sum_{m^\prime=0}^m \theta(1-\theta)^{m-m^\prime} p_A(m^\prime,t),
\end{split}
\end{align}
where $p_R(m,t)$ is the probability of the system having $m$ mRNA copies
and having repressor bound to the promoter at time $t$, and $p_A$ is an
analogous probability to find the promoter without repressor bound.
$k_R+$ and $k_R^-$ are, respectively, the rates at which repressors
bind and unbind to and from the promoter, and
$\gamma$ is the mRNA degradation rate.
$k_i$ is the rate at which bursts initiate, and as before,
the geometric distribution of burst sizes has mean $b=(1-\theta)/\theta$.\footnote{
$\theta$ can be viewed as a Bernoulli trial probability: after each transcript
in a burst is produced, $\theta$ is the probability that the burst
terminates and $(1-\theta)$ is the probability that the burst continues.
}

@@

any standard source, e.g.~\cite{Abramowitz1964}, and identify the

@@

\subsection{Numerical considerations and recursion formulas}
\subsubsection{Generalities}
We would like to carry out Bayesian parameter inference on FISH data
from~\cite{Jones2014}, using~\eref{eq:p_m_bursty+rep} as our
likelihood. This requires accurate (and preferably fast)
numerical evaluation of the hypergeometric function ${_2F_1}$,
which is a notoriously hard problem~\cite{Pearson2017, Gil2007},
and our particular needs here present an especial challenge as we show below.

@@

One solution is to evaluate ${_2F_1}$ using arbitrary precision arithmetic
instead of floating point arithmetic,
e.g., using the \texttt{mpmath} library in Python.
This is accurate but incredibly slow computationally.
To quantify how slow, we found that
evaluating the likelihood defined by~\eref{eq:p_m_bursty+rep} $\sim50$ times
(for a typical dataset of interest from~\cite{Jones2014},
with $m$ values spanning 0 to $\sim50$)
using arbitrary precision arithmetic is 100-1000 fold slower than
evaluating a negative binomial likelihood for the corresponding
constitutive promoter dataset.

To claw back $\gtrsim30$ fold of that slowdown, we can exploit
one of the many catalogued symmetries involving ${_2F_1}$.
The solution involves recursion relations originally explored by Gauss,
and studied extensively in~\cite{Pearson2017, Gil2007}.
They are sometimes known as contiguous relations and relate the values
of any set of 3 hypergeometric functions whose arguments differ by integers.
To rephrase this symbolically, consider a set of hypergeometric functions
indexed by an integer $n$,
\begin{align}
f_n = {_2F_1}(a+\epsilon_i n, b+\epsilon_j n, c+\epsilon_k n; z),
\end{align}
for a fixed choice of $\epsilon_i, \epsilon_j, \epsilon_k \in \{0,\pm 1\}$
(at least one of $\epsilon_i, \epsilon_j, \epsilon_k$ must be nonzero,
else the set of $f_n$ would contain only a single element).
Then there exist known recurrence relations of the form
\begin{align}
A_n f_{n-1} + B_n f_{n} + C_n f_{n+1} = 0,
\end{align}
where $A_n, B_n$, and $C_n$ are some functions of $a,b,c$, and $z$.
In other words, for fixed $\epsilon_i, \epsilon_j, \epsilon_k, a, b,$ and $c$,
if we can merely evaluate ${_2F_1}$ twice, say for $n^\prime$ and $n^\prime-1$,
then we can easily and rapidly generate values for arbitrary $n$.

This provides a convenient solution for our problem: we need repeated
evaluations of ${_2F_1}(a+m, b+m, c+m; z)$
for fixed $a,b$, and $c$ and many integer values of $m$.
They idea is that we can use arbitrary precision arithmetic to evaluate
${_2F_1}$ for just two particular values of $m$ and then generate
${_2F_1}$ for the other 50-100 values of $m$ using the recurrence
relation.\footnote{
There are even more sophisticated ways of utilizing the recurrence
relations that might have netted another factor of 2 speed-up, and
possibly as much as a factor of 10, but the method described here had
already reduced the computation time to an acceptable
$\mathcal{O}(\text{1 min})$, so these more sophisticated approaches did
not seem worth the time to pursue.
}

However, there two further wrinkles.
The first is that
a naive application of the recurrence relation is numerically unstable.
Roughly, this is because the three term recurrence relations,
like second order ODEs, admit two linearly independent solutions.
In a certain eigenbasis, one of these solutions dominates the other
as $n\rightarrow\infty$, and as $n\rightarrow-\infty$,
the dominance is reversed.
If we fail to work in this eigenbasis, our solution of the recurrence relation
will be a mixture of these solutions and rapidly accumulate numerical error.
For our purposes, it suffices to know that the authors of~\cite{Gil2007}
derived the numerically stable solutions (so-called \textit{minimal solutions})
for several possible choices of $\epsilon_i, \epsilon_j, \epsilon_k$.
Running the recurrence in the proper direction using a minimal solution
is numerically robust and can be done entirely in floating point arithmetic, 
so that we only need to evaluate ${_2F_1}$ with arbitrary precision arithmetic
to generate the seed values for the recursion.

The second wrinkle is a corollary to the first.
The minimal solutions are only minimal for certain ranges of the argument $z$,
and not all of the 26 possible recurrence relations
have minimal solutions for all $z$.
This can be solved by using one of the many transformation formulae for
${_2F_1}$ to convert to a different recurrence relation that has
a minimal solution over the required domain of $z$, although
this can require some trial and error to find the right transformation,
the right recurrence relation, and the right minimal solution.

\subsubsection{Particulars}
Let us now demonstrate these generalities for our problem of interest.
In order to evaluate the probability distribution of our
model,~\eref{eq:p_m_bursty+rep}, we need to evaluate hypergeometric functions
of the form ${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$
for values of $m$ ranging from $0$ to $\mathcal{O}(100)$.
The authors of~\cite{Gil2007} did not derive a recursion relation
for precisely this case. We could follow their methods and do so ourselves,
but it is much easier to convert to a case that they did consider.
The strategy is to look through the minimal solutions tabulated
in~\cite{Gil2007} and search for a transformation we could apply to
${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$ that would place the $m$'s
(the variable being incremented by the recursion)
in the same arguments of ${_2F_1}$ as the minimal solution.
After some ``guess and check,'' we found that the transformation
\begin{align}
{_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)
=
(1+b)^{-\alpha-m}
        {_2F_1}\left(\alpha+m, \gamma-\beta, \gamma+m; \frac{b}{1+b}\right),
\label{eq:rec_euler_pretransform}
\end{align}
produces a ${_2F_1}$ on the right hand side that closely resembles
the minimal solutions $y_{3,m}$ and $y_{4,m}$ in Eq.~4.3 in~\cite{Gil2007}.
Explicitly, these solutions are
\begin{align}
y_{3,m}
&\propto
{_2F_1}\left(-\alpha^\prime + \gamma^\prime - m,
                -\beta^\prime + \gamma^\prime,
                1-\alpha^\prime-\beta^\prime+\gamma^\prime-m;
                1-z\right)
\\
y_{4,m}
&\propto
{_2F_1}\left(\alpha^\prime + m,
                \beta^\prime,
                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
                1-z\right),
\label{eq:minimal_soln_sans_prefac}
\end{align}
where we have omitted prefactors which are unimportant for now.
Which of these two we should use depends on what values $z$ takes on.
Equating $1-z=b/(1+b)$ gives $z=1/(1+b)$, and since $b$ is strictly positive,
$z$ is bounded between 0 and 1.
From Eq.~4.5 in~\cite{Gil2007}, $y_{4,m}$ is the minimal solution
for real $z$ satisfying $0<z<2$, so this is the only minimal solution we need.

Now that we have our minimal solution,
what recurrence relation does it satisfy?
Confusingly, the recurrence relation of which $y_{4,m}$ is a solution
increments different arguments of ${_2F_1}$ that does $y_{4,m}$:
it increments the first only, rather than first and third.
This recurrence relation can be looked up, e.g., Eq.~15.2.10
in~\cite{Abramowitz1964}, which is
\begin{align}
(\gamma^\prime - (\alpha^\prime + m)) f_{m-1}
+
(2(\alpha^\prime+m) - \gamma^\prime + (\beta^\prime - \alpha^\prime)z)f_m
+ \alpha^\prime(z-1) f_{m+1} = 0.
\label{eq:chosen_rec_rel}
\end{align}
Now we must solve for the parameters appearing in the recurrence relation
in terms of our parameters, namely by setting
\begin{align}
\begin{split}
\alpha^\prime &= \alpha
\\
\beta^\prime &= \gamma - \beta
\\
1 + \alpha^\prime + \beta^\prime - \gamma^\prime &= \gamma
\\
1 - z &= \frac{b}{1+b}
\end{split}
\end{align}
and solving to find
\begin{align}
\begin{split}
\alpha^\prime &= \alpha
\\
\beta^\prime &= \gamma - \beta
\\
\gamma^\prime &= 1 + \alpha - \beta
\\
z &= \frac{1}{1+b}
.
\end{split}
\end{align}
Finally we have everything we need. The minimal solution
\begin{align}
y_{4,m}
=
\frac{\Gamma(1+\alpha^\prime-\gamma^\prime+m)}
        {\Gamma(1+\alpha^\prime+\beta^\prime-\gamma^\prime+m)}
\times
{_2F_1}\left(\alpha^\prime + m,
                \beta^\prime,
                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
                1-z\right),
\end{align}
where we have now included the necessary prefactors,
is a numerically stable solution of the recurrence
relation~\eref{eq:chosen_rec_rel} if the recursion is run
from large $m$ to small $m$.
So the procedure is to compute the value of ${_2F_1}$ for the two
largest $m$ values of interest using arbitrary precision arithmetic,
compute the prefactors to construct
$y_{4,\text{max}(m)}$ and $y_{4,\text{max}(m)-1}$,
recursively compute $y_{4,m}$ for all $m$ less than $\text{max}(m)$ down
to $m=0$, then cancel off the prefactors of the resulting values of
$y_{4,m}$ for all $m$ to produce ${_2F_1}$ for all desired $m$ values.
\mmnote{Maybe I should reformat this paragraph into
a pseudocode version of the algorithm?
Instead of, or in addition to?}

With ${_2F_1}$ computed, the only remaining numerical danger in computing
$p(m)$ in~\eref{eq:p_m_bursty+rep} is overflow of the gamma functions.
This is easily solved by taking the log of the entire expression
and using standard routines to compute the log of the gamma functions,
then exponentiating the entire expression at the end if $p(m)$
is needed rather than $\log p(m)$.

@@

% \bibliographystyle{nature}
\bibliographystyle{abbrv}
\bibliography{./library}
